# Reproducible Research: Peer Assessment 1

Martin Hediger
==============


## Loading and preprocessing the data

### 1. Loading the data
The data is loaded using

```{r cache=TRUE}
dat <- read.csv("activity.csv", header=TRUE)
```

### 2. Preprocessing the data
Using `head(dat)` and `str(dat)` the data is inspected.
Since the data appears tidy, no additional processing is carried out.

## What is mean total number of steps taken per day?

### 1 Histogram of the total number of steps taken each day
Using `tapply`, the data is factorized by date and then the `sum` function is applied to each cluster of data.
`NA` values are not removed from the data at this point because they would be converted to and enter the histogram as zeros.
Missing values are removed however when calculating the mean and median below.
The histogram of the so calculated daily total steps looks as follows.

```{r}
hist(tapply(dat$steps, dat$date, sum), breaks=10,
     main="Daily Total Step Distribution", xlab="Daily Total Steps", 
)
```

### 2 Mean and median total number of steps taken per day
To illustrate, the number of steps on a specific day can be calculated by:

```{r}
sum(dat[dat$date=="2012-11-29", 1])
```

Calculating the number of steps on each day can be done by calling (`NA` values are considered below):

```{r, results="hide"}
tapply(dat$steps, dat$date, sum)
```

The result of this calculation can also be obtained as a list (with appropriately labeled columns) by using `melt`:

```{r}
library(reshape2)
daily_sum <- melt(tapply(dat$steps, dat$date, sum))
colnames(daily_sum) <- c("Date", "Daily_Sum")
head(daily_sum)
```

The mean total numer of steps is then (removing `NA` values):

```{r}
mean(daily_sum$Daily_Sum, na.rm=TRUE)
```

Similary, the median over all steps over the whole observation period is then

```{r}
median(daily_sum$Daily_Sum, na.rm=TRUE)
```




## What is the average daily activity pattern?

### 1 Time series plot of 5-minute interval and average number of steps taken

In the plot, the *x*-axis are just the interval values and the *y*-axis values are the averages of the steps taken during each day specific interval point.
First, the values for the x-axis are collect. These can be arbitrarily from any day (since all days ran throught the same interval) and are subsequently converted to POSIX format.

```{r}
x <- dat[dat$date=="2012-10-01", 3]
xs <- sprintf("%04d", x)
xt <- strptime(xs, format="%H%M")
```

Then the values for the y-axis are collected.
Here, the aggregate function can be used to calculate the mean over all days at a particular interval-value (`agsteps`)

```{r, fig.width=10}
agsteps <- aggregate(. ~ interval, data=dat, FUN=mean)[2]
plot(xt, agsteps[[1]], type="l", xlab="Day Time", ylab="Aggregated Steps")
```

### 2 Maximum number of steps in 5-minute interval

```{r}
max(agsteps)
maxinterval <- which(agsteps==max(agsteps))
xt[maxinterval]
```

The average maximum number of steps (206) are within the interval at 08:35 in the morning.



## Imputing missing values

### Total number of missing values

A separate data frame consisting only of the data where `steps` are missing is constructed and by using the fact that R considers a `TRUE` value to equal 1, the number of `NA` values is calculated using `sum`:

```{r}
nas <- dat[is.na(dat$steps), ]
sum(is.na(nas$steps))
```

### Strategy for filling in missing values































## Are there differences in activity patterns between weekdays and weekends?

